A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-10 podc_paper/algorithm.tex
shortening

Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
Thus, the direct subblocks of $v.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
v.\fld{left.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{left}}+1..v.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
v.\fld{right.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{right}}+1..v.\fld{blocks}[b].\fld{end\sub{right}}].
\end{eqnarray}
This allows us to navigate to a block's direct subblocks easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
These fields allow us to pinpoint the  location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue $E_r$ in the linearization.
We do a binary search on \fld{sum\sub{enq}} fields of the root's blocks 
to locate the block that contains $E_r$.
If we know which block $B$ in some \node\ $v$ contains $E_r$,
we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains $E_r$
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
$E_r$.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
All information needed for this search process can be derived from the 
\eleft, \eright\ and \fld{sum\sub{enq}} fields.

@@

We now consider task \ref{findrank}.
First, we must determine whether the queue is empty when a dequeue occurs.
To facilitate this, each block in the root stores a \fld{size} field that stores the number of elements
in the queue after all operations in the linearization ordering up to that block (inclusive) 
have been performed.
We can easily determine which dequeues in a block $B_d$ in the root are null dequeues using
$B_{d-1}.\fld{size}$, which stores the size of the queue just before $B_d$'s operations are performed, and the number of enqueues and dequeues in $B_d$.
We can also compute the number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
as $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
Then, for any non-null dequeue in $B_d$, we can use this information to determine its
rank among all non-null dequeues in the linearization ordering, and hence the rank of the enqueue
whose value it should return (among all enqueues).

@@

Since the index of \var{B}'s superblock $B'$ is not known until $B$ is
propagated to $par$, $R\sub{\var{v}}$ cannot set the \nf{super} field of ${B}$ when
it creates \var{B}. 
It would be too costly to have $R_{par}$ set the \fld{super} field of all the direct subblocks of $B'$ (since
there could be up to $p$ direct subblocks).
Intuitively, the index of $B'$ will be close to the value of \var{par}.\fld{head} has when \var{B} is installed.
Since we can search for the precise superblock when we need it, it suffices that \var{B}.\fld{super}
is close to the actual index of  $B'$.
Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
We shall show \nf{$B$.\fld{super}}  differs
from the index of $B'$ by at most~1.

>>>>

Each block in an internal \node\ $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  
Thus, the direct subblocks of $v.\fld{blocks}[b]$ are
\begin{eqnarray}\label{defsubblock}
v.\fld{left.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{left}}+1..v.\fld{blocks}[b].\fld{end\sub{left}}] \mbox{ and}\nonumber\\
v.\fld{right.blocks}[v.\fld{blocks}[b-1].\fld{end\sub{right}}+1..v.\fld{blocks}[b].\fld{end\sub{right}}].
\end{eqnarray}
This allows us to navigate to a block's direct subblocks easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $\var{v}.\fld{blocks}[b]$ has two fields \fld{sum\sub{enq}} and \fld{sum\sub{deq}}
that store the total number of enqueues and dequeues in $\var{v}.\fld{blocks}[1..\var{b}]$.
These fields allow us to pinpoint the  location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue $E_r$ in the linearization.
A binary search for $r$ on \fld{sum\sub{enq}} fields of the root's blocks 
finds the block  containing $E_r$.
If we know a block $B$ in a \node\ $v$ contains $E_r$,
we can use the \fld{sum\sub{enq}} field again to determine which child of $v$ contains $E_r$
and then do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
$E_r$.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other \node\ along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps for task \ref{findenqueue}.
All information needed for this search process can be derived from the 
\eleft, \eright\ and \fld{sum\sub{enq}} fields.

@@

Now consider task \ref{findrank}.
To determine whether the queue is empty when a dequeue occurs,
each block in the root has a \fld{size} field storing the number of elements
in the queue after all operations in the linearization up to that block (inclusive) 
have been done.
We can  determine which dequeues in a block $B_d$ in the root are null dequeues using
$B_{d-1}.\fld{size}$, which is the size of the queue just before $B_d$'s operations, and the number of enqueues and dequeues in $B_d$.
Moreover, the total number of non-null dequeues in blocks $B_1, \ldots, B_{d-1}$ 
is $B_{d-1}.\fld{sum\sub{enq}}-B_{d-1}.\fld{size}$.
We can use this information to determine the
rank of a non-null dequeue in $B_d$ among all non-null dequeues in the linearization, which is the rank  (among all enqueues) of the enqueue
whose value the dequeue should return.

@@

Since the index of \var{B}'s superblock $B'$ is not known until $B$ is
propagated to $par$, $R\sub{\var{v}}$ cannot set the \nf{super} field of ${B}$ when
it creates \var{B}. 
It would be too costly to have $R_{par}$ set the \fld{super} field of all the direct subblocks of $B'$ (since
there could be up to $p$ direct subblocks).
Soon after $B$ is installed,
a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by some process will
set \var{B}.\fld{super} to a value read from \var{par}.\fld{head}.
We shall show that this happens soon enough that $B.\fld{super}$ differs from the true index of $B'$
by at most 1.
%Thus, after \var{B} is installed, the call to \op{Advance} at line \ref{advance} sets the value of \var{B}.\fld{super} by reading \var{par}.\fld{head}.
%As mentioned earlier, other processes may help by calling \op{Advance} at line \ref{helpAdvance} or \ref{advance} to ensure that $\var{B}.\fld{super}$ is filled in soon after \var{B} is installed.
%We shall show \nf{$B$.\fld{super}}  differs
%from the index of $B'$ by at most~1.

### 2023-01-10 podc_paper/analysis.tex
minor edits
shortening

\begin{proposition}
Each \nf{Enqueue} or \nf{Dequeue} operation performs $O(\log p)$ \nf{CAS} instructions.
\end{proposition}
\begin{proof}
An operation invokes \nf{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
A \nf{Refresh} does at most 7 \nf{CAS} steps: one in line \ref{cas} and two from each \nf{Advance} in line \ref{helpAdvance} or~\ref{advance}.
\end{proof}

@@

We now bound operations' running times in terms of the number of processes $p$, the maximum contention $c\leq p$ and the size of the queue.

\begin{lemma}\label{enqDeqTime}
Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{lemma}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
at each node on the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node on this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

>>>>

\begin{proposition}\label{casbound}
Each \nf{Enqueue} or \nf{Dequeue} operation performs $O(\log p)$ \nf{CAS} instructions.
\end{proposition}
\begin{proof}
An operation invokes \nf{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
A \nf{Refresh} does at most 7 \nf{CAS} steps: one in line \ref{cas} and two from each \nf{Advance} in line \ref{helpAdvance} or~\ref{advance}.
\end{proof}

@@

We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$ and the size of the queue.

\begin{mytheorem}\label{enqDeqTime}
Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{mytheorem}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
at each node on the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node on this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

### 2023-01-10 podc_paper/boundedspace.tex
shortening

Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{sec::GCanalysis} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(G)$ steps is spent on a GC phase.
There are at least $G$ operations between two GC phases,  so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

>>>>

A GC phase on a node takes $O(p\log p \log(p+q))$ steps to help complete all pending operations.
If all processes carry out this GC phase, it takes a total of $O(p^2\log p\log(p+q))=O(G\log(p+q))$ steps.
There are at least $G$ operations between two GC phases,  so for each node,
GC  contributes only $O(\log(p+q))$ steps to each operation in an amortized sense.
Adding up over all nodes an operation visits, 
an operation takes $O(\log p\log(p+q))$ steps doing GC in an amortized sense.
%Doing GC less frequently would yield better constant factors in the amortized number of steps per operation but use more space.

### 2023-01-10 podc_paper/conclusion.tex
shortening

We believe the approach used here to implement a lock-free queue 
could be applied to obtain other lock-free
data structures with a polylogarithmic number of steps per operation.
For example, we could implement a  vector data structure that stores a sequence and
provides three operations: \opa{append}{e} to add an element \var{e} to the end of the sequence,
\opa{get}{i} to read the \var{i}th element in the sequence, and
\opa{index}{e} to compute the position of element \var{e} in the sequence.
\here{Can omit rest of this paragraph to save space}
An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
\here{doublecheck all preceding claims for accuracy.}
We would like to investigate whether a similar approach could be used for stacks, deques or even priority queues.

>>>>

We believe the approach used here to implement a lock-free queue 
could be applied to obtain other lock-free
data structures with a polylogarithmic number of steps per operation.
For example, we could implement a  vector data structure that stores a sequence and
provides three operations: \opa{append}{e} to add an element \var{e} to the end of the sequence,
\opa{get}{i} to read the \var{i}th element in the sequence, and
\opa{index}{e} to compute the position of element \var{e} in the sequence.
\here{Can omit rest of this paragraph to save space}
%An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
%A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
%An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
\here{doublecheck all preceding claims for accuracy.}
We would like to investigate whether a similar approach could be used for stacks, deques or even priority queues.

### 2023-01-10 podc_paper/correctness.tex
shortening

After proving some basic invariants in Section \ref{sec::basicProperties},
we show in Section \ref{sec::propagating} that performing a double refresh at each node
is sufficient to propagate an operation to the root.
In Section \ref{sec::tracingCorrect} we show that the \op{getEnqueue} and \op{indexDequeue}
routines correctly navigate through the \ordering\ tree.
Finally, we show that the implementation is linearizable in Section \ref{sec::linearizability}.

>>>>

After proving some basic properties in Section \ref{sec::basicProperties},
we show in Section \ref{sec::propagating} that a double refresh at each node
suffices to propagate an operation to the root.
In Section \ref{sec::tracingCorrect} we show \op{GetEnqueue} and \op{IndexDequeue}
correctly navigate through the  tree.
Finally, we prove linearizability in Section \ref{sec::linearizability}.
